---
title: "Data Manipulation"
author: "Franco Sanchez"
date: "November 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Manipulation Assignment
<br>

### Using dplyr functions for data manipulation

`dyplyr` is an R Package that is meant to facilitate data manipulation. It does this by reducing the amount of syntax in the executable R code, as well as providing very efficient functions.
The major `dplyr` functions include: `select()`, `filter()`, `mutate()`, `group_by()`, and `summarize()`. The following exercise demonstrate how useful these functions are in data manipulation.

### Pipe and functions

`dplyr` includes a way to run multiple functions at once on a particular dataset. This is known as a "pipe", and it the code is `%>$`. For example, we use `select()` to focus on a partiuclar column of the data we are working with, and then `filter()` to extract partiuclar types of data.

```{r echo = FALSE}
survey <- read.csv("D:/Google Drive/Undergraduate Courses CSUMB/Fall 2016, Semester 7/STAT 210 Statistical Computing Seminar/R Markdown/Assignments/surveys.csv")
```

```{r message = FALSE, warning = FALSE}
## First we load the load the dplyr package
library(dplyr)

```
<br>
We can now apply the `dplyr` functions
```{r}
## Here we apply the select function to our survey data, and create a new dataframe with the variables of interests
survey1 <- select(survey, plot_id, species_id, weight)

## Using 'pipes' to apply multiple dplyr functions at once
survey1 <- survey %>%
            filter(weight < 5) %>% #we are interested all weight less than 5
            select(species_id, sex, weight) #without losing our previous modification, we move on to the next
head(survey1, n = 5)
```
<br>

######Challenge 1: Using pipes, subset the data to include individuals collected before 1995, and retain the columns year, sex, and weight.

```{r}
survey_1995 <- survey %>% # creating a new data set
  filter((year < 1995),#we extract all information prior to the year 1995
         !is.na(weight)) %>% #exclude missing variables in weight
  select(year, sex, weight) # along with the years, we want to include sex and weight

head(survey_1995, n = 5)
```
<br>

Next, we will apply the `mutate()` function, which generates new variables that are defined by the parameters we set
```{r}
survey %>%
  filter(!is.na(weight)) %>% ## exlcuding the missing weight values
  mutate(weight_kg = weight/1000) %>% ## This means we are creating a new column defined by the weight data divided by 1000 units
  head(n = 5) ## regular commands also work with dplyr pipes
```
<br>

Challenge 2: Create a new dataframe from the survey data that meets the following criteria: contains only the species\_id column and a column that contains values that are half the hindfoot\_length values. In this hindfoot\_half column, there are no NA values and all values are < 30.
```{r}
survey3 <- survey %>%
  filter(!is.na(hindfoot_length)) %>% #filtering out all NA values from hindfoot_length
  mutate(hindfoot_length_half = hindfoot_length / 2) %>% #defining a new variable hindfoot_length_half 
  filter(hindfoot_length_half < 30) %>% #filtering out all hindfoot_length_half values greater than 30  
  select(species_id, hindfoot_length_half) %>% #selcting these variables to be a part of the new dataframe
  head(n = 5)
  
range(survey3$hindfoot_length_half) #ensuring length is not over 30
```
<br>

Next we use the "split-apply-combine" data technique. This is usually done by using the `group_by()` and `summarize()` functions in tandem.
```{r}
##using the 'group' and 'summarize' functions
survey %>%
  filter(!is.na(weight), #get rid of all the NA values in weight
         sex != "") %>% #remove empty sex values in data
  group_by(sex, species_id) %>% #similar to select, we are only working with these two columns
  summarize(mean_weight = mean(weight), #we can analyze two or more variables at once, with each group obtaining a defined parameter
            min_weight = min(weight)) %>%
  print(n = 5)
```
<br>

```{r echo = FALSE}
survey_combined <- read.csv("D:/Google Drive/Undergraduate Courses CSUMB/Fall 2016, Semester 7/STAT 210 Statistical Computing Seminar/R Markdown/Assignments/combined.csv")
```

#### Final Challenges:
```{r}
#Challenge: How many individuals were caught in each plot_type surveyed?
survey_combined %>%
  select(plot_type) %>%
  tally() #tally allows us to determine the total observations of each variable level

#Challenge: find the mean, min, and max hindfoot length for each species 
survey_combined %>%
  group_by(species_id) %>%
  filter(!is.na(hindfoot_length)) %>%
  summarize(mean_hindfoot = mean(hindfoot_length),
            max_hindfoot = max(hindfoot_length),
            min_hindfoot = min(hindfoot_length)) %>%
  arrange(species_id, max_hindfoot, min_hindfoot, mean_hindfoot) %>% # arrange allows us to determine the order of columns in our dataset
  head(n = 5)

#Challenge: What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and  weight.
survey_combined %>%
  select(year, genus, species_id, weight) %>%
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  top_n(n = 1, wt = weight) %>% # this generates the top value(s) in 'n' columns, with the variable of interests being weight; if there are any ties, additonal columns will be made display tied values
  arrange(year, genus, species_id, weight) %>% #by default, arrange is done in ascending order
  print(n = 5)
```
<br>

### Exporting datasets
We can extract a modified dataset using the `write.csv()` command.
```{r eval = FALSE}

## Now we will clean the survey data and export it as a new CSV file
survey_clean <- survey_combined %>%
  filter(!is.na(weight), ##removing missing numeric values
         !is.na(hindfoot_length), 
         species_id != "", #removing all empty strings
         sex != "") 
         
## Now we will extract only the common species (exclude the rare ones)
species_count <- survey_clean %>% 
  group_by(species_id) %>% 
  tally %>% 
  filter(n >= 50) %>% ##filtering all species that are greater than or equal to 50 observations
  select(species_id)
  
## We will update our clean survey with this new info
survey_clean <- survey_clean %>%
  filter(species_id %in% species_count$species_id) ##we replace the original species_id with the updated one

## Let us export the new data as a new file, in a seperate folder from the original file to avoid overriding issues
write.csv(survey_clean, file = "survey_clean/surveys_clean.csv",
          row.names = FALSE) ## row names add numbered row names; we don't need them in this case
```