---
title: "Second R Markdown Assignment"
author: Franco Sanchez
date: September 16, 2016
output: html_document
---

```{r setup, include = FALSE}
    #Just exploring the data
data = read.csv(file.choose())
head(data)
summary(data)
str(data)

color <- data$color
summary(color)
factor(color) # This does the same thing as line 12
color_levels <- factor(color, order = FALSE, levels = c("White", "Pink", "Red"))
summary(color_levels) #THis does the same thing as line 10. so factor vector is not necessary
table(color)


#Creating a table of observation summary
color_summary <- summary(color_levels)
column_names <- c("Color", "Observations")
colors <- c("White", "Pink", "Red")
col_obs <- c("40", "34", "108")
table_summary <- c( colors, col_obs)
table_matrix <- matrix(data = table_summary, ncol = 2, byrow = FALSE)
colnames(table_matrix) <- c("Color", "Observations")

#Creating Figures with Data
barplot(prop.table(table(color)), xlab = "Colors", ylab = "Frequency", main = "Observation Frequency of Pink, Red and White")
barplot(table(color), xlab = "Colors", ylab = "Frequency", main = "Observation Count of Pink, Red, and White")
boxplot(summary(color), ylab = "All Colors: White, Pink, and Red", xlab = "Total Observations", main = "Boxplot of Color Data", horizontal = TRUE)
```


#Anlaysis of a Simple Dataframe

The data provided in "Poinsettias" is quite simple, with only two variables avaible for analysis (technically one since we ignore "id"). The important summary from this data is from the colors white, pink and red. The following table summarizes the observations:

```{r, echo = TRUE}
table_matrix
```

Here are some additonal statistic figures to accompany the table summary:

```{r, echo = FALSE}
barplot(prop.table(table(color)), xlab = "Colors", ylab = "Frequency", main = "Observation Frequency of Pink, Red and White")
barplot(table(color), xlab = "Colors", ylab = "Frequency", main = "Observation Count of Pink, Red, and White")
boxplot(summary(color), ylab = "All Colors: White, Pink, and Red", xlab = "Total Observations", main = "Boxplot of Color Data", horizontal = TRUE)
```

```{r, echo = FALSE}
col_obs <- c(40, 34, 108)
```

Finally, here we have the boxplot values (summary statistics):
```{r, echo = TRUE}
summary(col_obs) #Color Observations

