---
title: "Function Assignment"
author: "Franco Sanchez"
date: "October 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Assignment:
###Create your Own Function

You are going to create a function in R.  I do not care what it does, but here are the minimum requirements.

1. It must have at least three arguments
2. One of the arguments must have a default value
3. It should be able to validate argument types allowed (`stopifnot`)



### My Code

```{r}
test_num <- function(x, int = 1) 
{
  stopifnot(is.numeric(x))
  if (int == 1)
  x <- as.integer(x)
  
  else if(int == 0)
    x <- x
  
  if (x == 0)
  {
    print("The integer you intered is 0")
  }
  
  if (x > 0)
  {
    print("The integer you entered is positive")
  }
  
  if (x < 0)
  {
    print("The integer you entered is negative")
  }
  return(x)
}
```


####Description

This function tests if a number is either negative, 0, or positive.

####Usage

Notice that the variable 'int' is set to 1 by default. This tells the function that the input value 'x' will automatically be converted into an integer. The user has the option of turning this off by setting 'int' to 0.

####Arguments

The function makes the following arguments:

1. Checks if the input is a numeric value. If it is not numeric, an error message will be returned.
2. Checks if 'int' is either 1 or 0. If 1, then the input is an integer. If 0, the input is left alone (meaning the input can be in fraction form).
3. Checks if the input is negative in "if (x < 0)".
4. Checks if the input is 0 in "if (x == 0)".
5. Checks if the input is positive in "if (x > 0)"

####Details

This is a simple function that can be implemented in other bigger functions that rely on the sing of a number. However, this function is breakable by several arguments that involve numeric values, such as a list or vector input. This function would require further "if - stop" arguments to prevent breaking.

####Examples

```{r}
test_num(-2)
test_num(0)
test_num(3.9)
test_num(3.9, 0)
```

